---
title: "Data science with `{hyenaR}`:<br>LESSON 10"
format:
  revealjs:
    code-overflow: scroll
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
#| code-line-numbers: "7"
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
library(stringr) ## Working text
library(waldo) ## To compare objects
```

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

------------------------------------------------------------------------

### GOAL 1: 📝Homework (Lesson 8)

<br>

### GOAL 3: `select()`/`where()`

<br>

### GOAL 2: `group_by()`/`filter()`

# 📝Homework (Lesson 8)

## [TASK 1:]{style="color:#ff2c2d;"} {auto-animate="true"}

### Use `create_id_starting.table()` to find all individuals born in main clans in the years 1997 - 2021.

```{r}
#| echo: true
#| code-line-numbers: "1|2|3|4"
cubs <- create_id_starting.table(lifestage = "cub",
                                 lifestage.overlap = "start",
                                 from = "1997-01-01", to = "2021-12-31",
                                 clan = find_clan_name.all(main = TRUE))
```

## [TASK 1:]{style="color:#ff2c2d;"} {auto-animate="true"}

### How many individuals have been born in total during this period?

```{r}
#| echo: true
nrow(cubs)
```

## [TASK 1:]{style="color:#ff2c2d;"} {auto-animate="true"}

::: callout-tip
## NOTE:

By itself, 'clan.birth' will *not* help here because it ignores from/to. 'clan.birth' is intended as a filter in *addition* to other arguments.
:::

```{r}
#| echo: true
create_id_starting.table(clan.birth = find_clan_name.all(main = TRUE),
                         from = "1997-01-01", to = "2021-12-31") %>% 
  nrow()
```

## [TASK 2:]{style="color:#ff2c2d;"} {auto-animate="true"}

### Extract individual birth clan, birth date, sex, and lifespan.

```{r}
#| echo: true
#| code-line-numbers: "2|3|4|5"
cubs_wdetails <- cubs %>% 
  mutate(birth_clan = fetch_id_clan.birth(ID),
         birth_date = fetch_id_date.birth(ID),
         sex = fetch_id_sex(ID),
         lifespan = fetch_id_duration.lifespan(ID))

cubs_wdetails
```

## [TASK 2:]{style="color:#ff2c2d;"} {auto-animate="true"}

::: callout-tip
## NOTE:

`fetch_id_duration.lifespan()` and `fetch_id_age(at = death_date)` are the same.
:::

```{r}
#| echo: true
waldo::compare(fetch_id_duration.lifespan(cubs$ID),
               fetch_id_age(cubs$ID, at = fetch_id_date.death(ID = cubs$ID)))
```

## [TASK 2:]{style="color:#ff2c2d;"} {auto-animate="true"}

### How many individuals have missing data?

```{r}
#| echo: true
cubs_wdetails %>% 
  filter(if_any(.cols = everything(),
                .fns = is.na)) %>% 
  nrow()
```

## [TASK 2:]{style="color:#ff2c2d;"} {auto-animate="true"}

### How would you interpet NAs in each column?

::: incremental
-   **birth clan/birth date**: A mistake in the data! All individuals should have an assigned birth clan and date.

-   **sex**: Sex not (yet) determined. Particularly common for individuals that died young.

-   **lifespan/death date**: Individual is right censored (i.e. hasn't died yet!)
:::

## [TASK 2:]{style="color:#ff2c2d;"} {auto-animate="true"}

### [BONUS:]{style="color:#ff2c2d;"} `select()`/`where()`

::: callout-tip
## NOTE:

If we are just applying one function we can just write the function name.
:::

```{r}
#| echo: true
#| code-line-numbers: "2-3"
cubs_wdetails %>%
  #Select columns that are numeric
  select(where(fn = is.numeric))
```

## [TASK 2:]{style="color:#ff2c2d;"} {auto-animate="true"}

### [BONUS:]{style="color:#ff2c2d;"} `select()`/`where()`

::: callout-tip
## NOTE:

If we want something more complicated we need to include our own function.
:::

```{r}
#| echo: true
#| code-line-numbers: "2-3"
cubs_wdetails %>%
  #Select columns that contain at least 1 NA
  select(where(fn = function(x) any(is.na(x))))
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### Extract the *year* of birth for each individual.

```{r}
#| echo: true
#| code-line-numbers: "2"
cubs_wbirthyr <- cubs_wdetails %>% 
  mutate(birth_year = lubridate::year(birth_date))

cubs_wbirthyr
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### In which year were the most cubs born?

```{r}
#| echo: true
#| code-line-numbers: "2|3|4|5"
cubs_wbirthyr %>% 
  group_by(birth_year) %>% 
  summarise(total_cubs = n()) %>% 
  ungroup() %>% 
  arrange(desc(total_cubs))
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### Is the most productive year the same for all clans?

```{r}
#| echo: true
#| code-line-numbers: "2|3|4|5"
cubs_wbirthyr %>% 
  group_by(birth_year, birth_clan) %>% 
  summarise(total_cubs = n()) %>% 
  ungroup() %>% 
  arrange(desc(total_cubs))
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### [BONUS:]{style="color:#ff2c2d;"} Use filter to return a data frame with only the best year(s) for each clan?

```{r}
#| echo: true
#| code-line-numbers: "5|6-7"
cubs_wbirthyr %>% 
  group_by(birth_year, birth_clan) %>% 
  summarise(total_cubs = n()) %>% 
  ungroup() %>% 
  group_by(birth_clan) %>% 
  #Return the record that has the max number of cubs in EACH CLAN
  filter(total_cubs == max(total_cubs))
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### [BONUS:]{style="color:#ff2c2d;"} `group_by()`/`filter()`

This allows us to apply filters to each group separately!

```{r}
#| echo: true
#| code-line-numbers: "2|3|4-5"
cubs_wbirthyr %>% 
  # Find longest lifespan in each clan!
  group_by(birth_clan) %>%
  # Find record that has max lifespan in EACH CLAN
  filter(lifespan == max(lifespan, na.rm = TRUE))
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### [BONUS:]{style="color:#ff2c2d;"} `group_by()`/`filter()`

Find earliest and latest individual(s) born in each year

```{r}
#| echo: true
#| code-fold: true
#| code-line-numbers: "2|3-4|5-6"
cubs_wbirthyr %>% 
  group_by(birth_year) %>% 
  # Apply two filters with |
  filter(birth_date == min(birth_date) | birth_date == max(birth_date)) %>% 
  ungroup() %>% 
  arrange(birth_year)
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

### [BONUS:]{style="color:#ff2c2d;"} `group_by()`/`filter()`

Find earliest and latest individual(s) born in Airstrip each year

```{r}
#| echo: true
#| code-fold: true
#| code-line-numbers: "2-3|4|5-6|7-8"
cubs_wbirthyr %>% 
  # We need to filter here FIRST...
  filter(birth_clan == "A") %>% 
  group_by(birth_year) %>% 
  #min/max are applied to the whole group
  filter(birth_date == min(birth_date) | birth_date == max(birth_date)) %>% 
  ungroup() %>% 
  arrange(birth_year)
```

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

::: callout-warning
This is why we always need to be careful to check grouping!!
:::

::: columns
::: {.column width="50%"}
### GROUPED

```{r}
#| echo: true
cubs_wbirthyr %>% 
  group_by(birth_year, birth_clan) %>% 
  summarise(total_cubs = n())
```
:::

::: {.column width="50%"}
### UNGROUPED

```{r}
#| echo: true
cubs_wbirthyr %>% 
  group_by(birth_year, birth_clan) %>% 
  summarise(total_cubs = n()) %>% 
  ungroup()
```
:::
:::

## [TASK 3:]{style="color:#ff2c2d;"} {auto-animate="true"}

::: callout-warning
This is why we always need to be careful to check grouping!!
:::

::: columns
::: {.column width="50%"}
### GROUPED

```{r}
#| echo: true
cubs_wbirthyr %>% 
  group_by(birth_year, birth_clan) %>% 
  summarise(total_cubs = n()) %>% 
  filter(total_cubs == max(total_cubs))
```
:::

::: {.column width="50%"}
### UNGROUPED

```{r}
#| echo: true
cubs_wbirthyr %>% 
  group_by(birth_year, birth_clan) %>% 
  summarise(total_cubs = n()) %>% 
  ungroup() %>% 
  filter(total_cubs == max(total_cubs))
```
:::
:::

## [TASK 4:]{style="color:#ff2c2d;"} {auto-animate="true"}

### Extract the *month* of birth for each individual.

```{r}
#| echo: true
#| code-line-numbers: "2"
cubs_wbirthyr_month <- cubs_wbirthyr %>% 
  mutate(birth_month = lubridate::month(birth_date))

cubs_wbirthyr_month
```

## [TASK 4:]{style="color:#ff2c2d;"} {auto-animate="true"}

### What was the most productive month in the crater from 1997-2021? How many cubs were produced in that month?

```{r}
#| echo: true
#| code-line-numbers: "2|3-4|5"
cubs_wbirthyr_month %>% 
  group_by(birth_year, birth_month) %>% 
  summarise(total_cubs = n()) %>% 
  ungroup() %>% 
  filter(total_cubs == max(total_cubs))
```

# 📝[HOMEWORK:]{style="color:#ff2c2d;"} Lesson 10.

## [TASK 1:]{style="color:#ff2c2d;"}

### Create a data frame of all females.

<br>

::: fragment
### Use `group_by()`/`summarise()` to show how many individuals are left or right censored.
:::

<br>

::: fragment
### Filter only un-censored individuals.
:::

## [TASK 2:]{style="color:#ff2c2d;"}

### Extract lifetime reproductive success of females (consider only offspring that reach atleast 18 months old).

<br>

::: fragment
### Extract standardised rank within females at the date of adulthood (i.e. when they were 2 years old).
:::

<br>

::: fragment
### Inspect the data. Are there any individuals you might exclude?
:::

## [TASK 3:]{style="color:#ff2c2d;"}

### Create a new column with rank category (top rank \>0 or low ranking \<=0). [HINT:]{style="color:#ff2c2d;"} You might need to use `ifelse`.

<br>

::: fragment
### What is the mean and SE of lifetime reproductive success in each group? [HINT:]{style="color:#ff2c2d;"} Remember SE = $\frac{sd}{\sqrt N}$
:::

<br>

::: fragment
### [BONUS:]{style="color:#ff2c2d;"} Look at the help for the `case_when()` function. Can you use this to make 3 groups instead of 2?
:::
