---
title: "Data science with `{hyenaR}`:<br>LESSON 17"
format:
  revealjs:
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# 🎉`hyenaR` v0.9.999941🎉

## Use `{drat}` to access the new version of `{hyenaR}`.

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1-2|4-5|7-8"
## To download package from other sources
library(drat)

## Include 'hyenaproject' as a package source
addRepo("hyenaproject") 

## Download hyenaR
install.packages("hyenaR")
```

<br>

::: fragment
```{r}
#| echo: true
#| code-line-numbers: "1-2"
#Check you have the right version (0.9.999941)
packageVersion("hyenaR")
```
:::

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
library(stringr) ## Working with text
library(waldo) ## To compare objects
library(skimr) ## Inspect data
library(purrr) ## For loops in the tidyverse
```

------------------------------------------------------------------------

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

------------------------------------------------------------------------

<br>

### GOAL 1: 🎉 Introduce the new features of v0.9.999941

<br>

### GOAL 2: 🧑‍🏫 Functions and debugging in R

# GOAL 1: Introduce the new features of v0.9.999941

## Check the NEWS to find out more

```{r}
#| eval: false
#| echo: true
build_vignette_news()
```

## Much faster! {auto-animate="true"}

```{r}
#| eval: false
#| echo: true
fetch_clan_number.anysex.all(clan = "A", at = "1997-01-01")
```

<br>

::: columns
::: {.column width="50%"}
::: {.fragment fragment-index="1"}
#### v0.9.99994
```{r}
#| eval: true
#| echo: false
print("Median time: ~1.04s")
```
:::
:::
::: {.column width="50%"}
::: {.fragment fragment-index="2"}
#### v0.9.999941
```{r}
#| eval: true
#| echo: false
print("Median time: ~387ms")
```
:::
:::
:::

<br><br>

::: {.fragment fragment-index="3"}
### ~3x FASTER!
:::

# GOAL 2: 🧑‍🏫 Functions and debugging in R

## Why create functions? {auto-animate="true"}

::: incremental
-   Reproducible, consistent behaviour.

-   Great for collaboration.

-   The 'Rule of Three'
:::

## Creating a function {auto-animate="true"}

```{r}
#| echo: true
#| eval: false
# Our function name...
my_function
```

## Creating a function {auto-animate="true"}

```{r}
#| echo: true
#| eval: false
# Our function name and argument names
my_function <- function(argument1, argument2){}
```

## Creating a function {auto-animate="true"}

```{r}
#| echo: true
# Our function name and argument names
my_function <- function(argument1, argument2){
  
  # Code between {}
  argument1 + argument2
  
}
```

::: {.fragment}
```{r}
#| echo: true
my_function(10, 1)
```
:::

## Using a function {auto-animate="true"}

You can specify the value of each argument explicitly
```{r}
#| echo: true
my_function(argument1 = 1, argument2 = 3)
```

::: {.fragment}
...or use the order or arguments.
```{r}
#| echo: true
my_function(1, 3)
```
:::

::: {.fragment}
Specifying argument names is often better to prevent bugs!
<br>
```{r}
#| echo: true
my_function(argument2 = 3, argument1 = 1)
```
:::

## Using a function {auto-animate="true"}

::: callout-note
Not all functions need arguments!
:::

```{r}
#| echo: true
day_of_the_week <- function(){
  format(Sys.Date(), "%A")
}
day_of_the_week()
```

## Default values {auto-animate="true"}

Using default values can make functions more flexible but still user friendly.

::: columns
::: {.column width="50%"}
::: {.fragment fragment-index="1"}
#### This function is inflexible
```{r}
#| echo: true
add <- function(number){
  
  number + 1
  
}
add(number = 10)
```
:::
:::
::: {.column width="50%"}
::: {.fragment fragment-index="2"}
#### This function is flexible but more complex
```{r}
#| echo: true
add <- function(number, addition){
  
  number + addition
  
}
add(number = 10, addition = 1)
```
:::
:::
:::

## Default values {auto-animate="true"}

Using default values can make functions easier to use, but still flexible.

<br>

#### This function is flexible but simple to use for the default case!
```{r}
#| echo: true
add <- function(number, addition = 1){
  
  number + addition
  
}
add(number = 10)
add(number = 10, addition = 2)
```

## What does a function return? {auto-animate="true"}

By default, a function returns the last value output by the code...

```{r}
#| echo: true
add <- function(number, addition = 1){
  
  number + addition
  10
  
}
```

What will this return?
```{r}
#| echo: true
#| eval: false
add(20)
```

::: {.fragment fragment-index="2"}
```{r}
#| echo: false
#| eval: true
add(20)
```
:::

## What does a function return? {auto-animate="true"}

Use `return()` if you want to be specific.

::: callout-warning
Everything after `return()` is not run!
:::

```{r}
#| echo: true
add <- function(number, addition = 1){
  
  new_number <- number + addition
  return(new_number)
  10
  
}
```

What will this return?
```{r}
#| echo: true
#| eval: false
add(20)
```

::: {.fragment fragment-index="2"}
```{r}
#| echo: false
#| eval: true
add(20)
```
:::

## What does a function return? {auto-animate="true"}

`return()` can be more efficient!

```{r}
#| echo: true
#| code-line-numbers: "1|2-5|6-9|10-13|14-15"
is_even <- function(object){
  ## Check if a number is 0...
  if (object == 0) {
    output <- NA
  }
  ## THEN check if a number is >0 and even
  if (object > 0 & object %% 2 == 0) {
    output <- "even"
  }
  ## THEN check if a number is >0 and odd
  if (object > 0 & object %% 2 != 0) {
    output <- "odd"
  }
  ## Return the output
  output
}
```

## What does a function return? {auto-animate="true"}

`return()` can be more efficient!

```{r}
#| echo: true
#| code-line-numbers: "2-5"
is_even_new <- function(object){
  ## If number is 0...return NA and stop!!
  if (object == 0) {
    return(NA)
  }
  ## Otherwise...check if a number is >0 and even
  if (object > 0 & object %% 2 == 0) {
    output <- "even"
  }
  ## and check if a number is >0 and odd
  if (object > 0 & object %% 2 != 0) {
    output <- "odd"
  }
  ## Return the output
  output
}
```

## What does a function return? {auto-animate="true"}

`return()` can be more efficient!

```{r}
#| echo: false
bench::mark(is_even(0), is_even_new(0)) %>% 
  dplyr::select(expression, median, `itr/sec`) %>% 
  dplyr::mutate(expression = as.character(expression))
```

## What if something goes wrong? {auto-animate="true"}

Default error messages are often hard to understand.

```{r}
#| echo: true
#| error: true
add <- function(number, addition = 1){
  
  number + addition
  
}
add("B")
```

## What if something goes wrong? {auto-animate="true"}

Use `stop()` to create your own error message!

```{r}
#| echo: true
#| error: true
add <- function(number, addition = 1){
  
  if (is.character(number) | is.character(addition)) {
    stop("The arguments 'number' and 'addition' cannot be characters!")
  }
  number + addition
  
}
add("B")
```

## What if something goes wrong? {auto-animate="true"}

Use `warning()` to flag things that can work but might cause problems.

```{r}
#| echo: true
#| warning: true
add <- function(number, addition = 1){
  
  if (is.character(number) | is.character(addition)) {
    stop("The arguments 'number' and 'addition' cannot be characters!")
  }
  
  if (is.logical(number) | is.logical(addition)) {
    warning("Coercing logical value to numeric.\n")
  }
  number + addition
  
}
add(number = 10, addition = TRUE)
```

## Debugging! {auto-animate="true"}

::: incremental
-   `browser()`

-   `debug()`

-   `trace()`
:::

## Debugging! {auto-animate="true"}

`browser()` can help debug a function you are creating!

```{r}
#| echo: true
#| error: true
day_of_week <- function(date){
  if (is.numeric(date)) {
    stop("Function cannot accept a number")
  }
  day <- format(as.Date(date), "%A")
  paste("You were born on", day)
}
day_of_week("12-03-88")
```

## Debugging! {auto-animate="true"}

The function will stop at the point where we add `browser()`...

```{r}
#| echo: true
#| eval: false
day_of_week <- function(date){
  browser()
  if (is.numeric(date)) {
    stop("Function cannot accept a number")
  }
  day <- format(as.Date(date), "%A")
  paste("You were born on", day)
}
day_of_week("12-03-88")
```

## Debugging! {auto-animate="true"}

Use the 'expr' argument in `browser()` to debug in certain conditions!

```{r}
#| echo: true
transform_numbers <- function(n){
  
  output <- c()
  
  for (i in runif(min = -10, max = 10, n = n)) {
    
    output <- append(output, log(i))
    
  }
  
  return(output)
  
}
```
```{r}
#| echo: true
set.seed(123)
transform_numbers(10)
```

## Debugging! {auto-animate="true"}

Use the 'expr' argument in `browser()` to debug in certain conditions!

```{r}
#| echo: true
transform_numbers <- function(n){
  
  output <- c()
  
  for (i in runif(min = -10, max = 10, n = n)) {
    
    browser(expr = i <= 0)
    output <- append(output, log(i))
    
  }
  
  return(output)
  
}
```
```{r}
#| echo: true
#| eval: false
set.seed(123)
transform_numbers(10)
```

## Debugging! {auto-animate="true"}

`debug()` can be used for any function (even ones you didn't write!)

```{r}
#| echo: true
#| eval: false
debug(day_of_week)
day_of_week("12-03-88")
```

::: callout-warning
You will need to `undebug()` a function (or use `debugonce()`)
:::

## Debugging! {auto-animate="true"}

If you identify a problem in another function, you can use `trace(edit = TRUE)` to tweak it.

```{r}
#| echo: true
date_time <- lubridate::ymd_hms("1999-01-01 00:00:00", tz = "Africa/Dar_es_Salaam")
as.Date(date_time)
```

## Debugging! {auto-animate="true"}

Try it yourself.

```{r}
#| echo: true
average_age_firstrepro <- function(sex, birth_year, na.rm = FALSE){
  
  ## Create start and end date for the target year
  start_date <- paste('birth_year', "01/01", sep = "_")
  end_date   <- paste('birth_year', "12/31", sep = "_")
  
  ## Extract all individuals born in that year and their age at first birth
  all_births <- create_id_starting.table(from = start_date, to = end_date,
                           sex = sex,
                           lifestage = "cub", lifestage.overlap = "start") %>% 
    dplyr::mutate(firstbirth = fetch_id_date.birth.first(ID)) %>% 
    dplyr::mutate(age = fetch_id_age(ID, at = firstbirth))
  
  ## Determine an average age at first birth
  avg_age <- mean(all_births$age, na.rm = na.rm)
  
  ## Return the number
  return(avg_age)

}
```

## Debugging! {auto-animate="true"}

Why doesn't this code work?

::: callout-note
There might be more than one problem...
:::

```{r}
#| echo: true
#| eval: false
average_age_firstrepro(2019, "male")
```

