---
title: "Data science with `{hyenaR}`:<br>LESSON 15"
format:
  revealjs:
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
library(stringr) ## Working with text
library(waldo) ## To compare objects
library(skimr) ## Inspect data
```

------------------------------------------------------------------------

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

# GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}`

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### Extract monthly female reproduction and clan size for Airstrip. Build a model to study the relationship.

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### [Extract monthly female reproduction]{style="color:#ff2c2d;"} and clan size for Airstrip. Build a model to study the relationship.

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### [Extract monthly female reproduction]{style="color:#ff2c2d;"} [and clan size for Airstrip]{style="color:#1b91ff;"}. Build a model to study the relationship.

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### [Extract monthly female reproduction]{style="color:#ff2c2d;"} [and clan size for Airstrip]{style="color:#1b91ff;"}. [Build a model to study the relationship.]{style="color:#fda703;"}

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "1-2|4-9"
start_date <- find_pop_date.observation.first()
end_date <- find_pop_date.observation.last()

## Extract all adult females in Airstrip during observation period
create_id_starting.table(sex = "female",
                         clan = "A",
                         lifestage = "adult",
                         from = start_date,
                         to = end_date)
```

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "5-7"
start_date <- find_pop_date.observation.first()
end_date <- find_pop_date.observation.last()

create_id_starting.table(sex = "female", clan = "A", lifestage = "adult", from = start_date, to = end_date) %>% 
  ## Extract dates over which individual could reproduce...
  mutate(adult_date = fetch_id_date.at.age(ID = ID, age = 2),
         death_date = fetch_id_date.death(ID = ID))
```

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "7|8-9|10-11"
start_date <- find_pop_date.observation.first()
end_date <- find_pop_date.observation.last()

create_id_starting.table(sex = "female", clan = "A", lifestage = "adult", from = start_date, to = end_date) %>% 
  mutate(adult_date = fetch_id_date.at.age(ID = ID, age = 2),
         death_date = fetch_id_date.death(ID = ID)) %>% 
  ## Use case when to deal with left and right censored individuals...
  mutate(start_date = case_when(adult_date < start_date ~ start_date,
                                TRUE ~ start_date),
         end_date = case_when(is.na(death_date) ~ end_date,
                              TRUE ~ death_date))
```

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "11|12|13"
start_date <- find_pop_date.observation.first()
end_date <- find_pop_date.observation.last()

create_id_starting.table(sex = "female", clan = "A", lifestage = "adult", from = start_date, to = end_date) %>% 
  mutate(adult_date = fetch_id_date.at.age(ID = ID, age = 2),
         death_date = fetch_id_date.death(ID = ID)) %>% 
  mutate(start_date = case_when(adult_date < start_date ~ start_date,
                                TRUE ~ start_date),
         end_date = case_when(is.na(death_date) ~ end_date,
                              TRUE ~ death_date)) %>% 
  ## Only use months where individual was alive the whole time
  mutate(start_date = lubridate::ceiling_date(start_date, unit = "month"),
         end_date = lubridate::floor_date(end_date, unit = "month"))
```

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "15-18"
start_date <- find_pop_date.observation.first()
end_date <- find_pop_date.observation.last()

ID_data <- create_id_starting.table(sex = "female", clan = "A", lifestage = "adult", from = start_date, to = end_date) %>% 
  mutate(adult_date = fetch_id_date.at.age(ID = ID, age = 2),
         death_date = fetch_id_date.death(ID = ID)) %>% 
  mutate(start_date = case_when(adult_date < start_date ~ start_date,
                                TRUE ~ start_date),
         end_date = case_when(is.na(death_date) ~ end_date,
                              TRUE ~ death_date)) %>% 
  mutate(start_date = lubridate::ceiling_date(start_date, unit = "month"),
         end_date = lubridate::floor_date(end_date, unit = "month")) %>% 
  ## Create one row per month of life...
  reshape_row_date.seq(ID, from = start_date, to = end_date, by = "1 month") %>% 
  group_by(ID) %>% 
  mutate(end_date = lead(date) - 1) %>% 
  ungroup() %>% 
  filter(!is.na(end_date))
```

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
ID_data
```

## [Extract monthly female reproduction...]{style="color:#ff2c2d;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "2-5"
(repro_data <- ID_data %>% 
  mutate(RS = fetch_id_number.offspring(ID, from = date, to = end_date,
                                        age.mature = 6,
                                        unit = "month"),
         repro = RS > 0))
```

## [...and clan size for Airstrip]{style="color:#1b91ff;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "1|2-8"
tibble(clan = "A") %>% 
  #Expand out for Airstrip over the same dates...
  reshape_row_date.seq(clan,
                       from = ceiling_date(start_date, unit = "month"),
                       to = floor_date(end_date, unit = "month"),
                       by = "month") %>% 
  mutate(end_date = lead(date) - 1) %>% 
  filter(!is.na(end_date))
```

## [...and clan size for Airstrip]{style="color:#1b91ff;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "5-6"
clan_data <- tibble(clan = "A") %>% 
  reshape_row_date.seq(clan, from = ceiling_date(start_date, unit = "month"), to = floor_date(end_date, unit = "month"), by = "month") %>% 
  mutate(end_date = lead(date) - 1) %>% 
  filter(!is.na(end_date)) %>% 
  mutate(clan_size = fetch_clan_number(clan = clan,
                                       from = date, to = end_date))
```

## [...and clan size for Airstrip]{style="color:#1b91ff;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
clan_data
```

## [...and clan size for Airstrip]{style="color:#1b91ff;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
final_data <- repro_data %>% 
  left_join(clan_data, by = c("date", "end_date"))
```

## [Build a model to study the relationship.]{style="color:#fda703;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
mod <- glm(repro ~ clan_size, data = final_data,
           family = "binomial")

summary(mod)
```

## [Build a model to study the relationship.]{style="color:#fda703;"}{auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-fold: true
#| fig-width: 7
#| fig-height: 5
pred_data <- data.frame(clan_size = seq(30, 140, 1)) %>% 
  mutate(pred = predict(mod, newdata = ., type = "response"))

plot_data <- final_data %>% 
  mutate(clan_size_10 = (clan_size %/% 10) * 10 + 5) %>% 
  group_by(clan_size_10) %>% 
  summarise(binom::binom.wilson(x = sum(repro), n = n()))

ggplot() +
  geom_errorbar(data = plot_data, aes(x = clan_size_10,
                                      ymin = lower, ymax = upper),
                width = 2.5) +
  geom_point(data = plot_data, aes(x = clan_size_10,
                                   y = mean),
             shape = 21, colour = "white", fill = "grey10",
             size = 2) +
  geom_line(data = pred_data, aes(x = clan_size, y = pred),
            linewidth = 1) +
  labs(y = "Probability of reproduction",
       x = "Clan size",
       title = "Airstrip clan") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.01)) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black", size = 10))
```

