---
title: "Data science with `{hyenaR}`:<br>LESSON 8"
format:
  revealjs:
    code-overflow: scroll
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# 🎉`hyenaR` v0.9.99994🎉

## Use `{drat}` to access the new version of `{hyenaR}`.

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1-2|4-5|7-8"
## To download package from other sources
library(drat)

## Include 'hyenaproject' as a package source
addRepo("hyenaproject") 

## Download hyenaR
install.packages("hyenaR")
```

<br>

::: fragment
```{r}
#| echo: true
#| code-line-numbers: "1-2"
#Check you have the right version (0.9.99994)
packageVersion("hyenaR")
```
:::

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
```

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

## GOAL 1: Introduce the new features of v0.9.99994

# Check the NEWS to find out more

```{r}
#| eval: false
#| echo: true
build_vignette_news()
```

# Relatedness functions

## 👪 New and old relatedness functions {auto-animate="true"}

```{r}
#| echo: true
#This is now *correct* relatedness using `kinship2`
fetch_dyad_relatedness(ID.1 = "T-212", ID.2 = "T-177")
```

<br>

::: fragment
```{r}
#| echo: true
#This is the previous (incorrect) relatedness used in social support
fetch_dyad_relatedness.via.filiation(ID.1 = "T-212", ID.2 = "T-177", filiation = "father")
```
:::

## 👪 New and old relatedness functions {auto-animate="true"}

```{r}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 7
#Find relatedness of all individuals in Triangle at a given date
Triangle_IDs <- find_clan_id(clan = "T", from = "2000-01-01", to = "2001-01-01")

#Estimate relatedness for the first 50
df <- expand.grid(ID.1 = Triangle_IDs,
                  ID.2 = Triangle_IDs) %>% 
  mutate(old_relatedness = fetch_dyad_relatedness.via.filiation(ID.1 = ID.1,
                                                                ID.2 = ID.2,
                                                                filiation = "mother_genetic", verbose = FALSE),
         new_relatedness = fetch_dyad_relatedness(ID.1 = ID.1,
                                                  ID.2 = ID.2)) %>% 
  #When there's no relatedness on mother genetic assume it's 0
  mutate(old_relatedness = tidyr::replace_na(old_relatedness, 0))

df %>% 
  mutate(diff = abs(old_relatedness - new_relatedness)) %>%
  {ggplot(.) +
      geom_histogram(aes(x = diff), binwidth = 0.01) +
      labs(title = "Functions will have given similar results (but not always)",
           subtitle = "Triangle clan member during the year 2000") +
      labs(x = "Difference between old and\nnew relatedness measures") +
      theme_classic() +
      theme(plot.title = element_text(face = "bold"))}
```

# create_id_starting.table()

## 🐛 Bug fix {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1|2-5"
# OLD BEHAVIOUR: Dispersers are skipped incorrectly
create_id_starting.table.historic(lifestage = "!dead",
                                  from = "1998-01-01",
                                  to = "1998-12-31",
                                  lifestage.overlap = "always", verbose = FALSE)
```

## 🐛 Bug fix {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1|2-5"
# NEW BEHAVIOUR: Dispersers are included!
create_id_starting.table(lifestage = "!dead",
                         from = "1998-01-01",
                         to = "1998-12-31",
                         lifestage.overlap = "always", verbose = FALSE)
```

## 🐛 Bug fix {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1-2"
# NEW BEHAVIOUR: 'alive' lifestage is possible
create_id_starting.table(lifestage = "alive",
                         from = "1998-01-01",
                         to = "1998-12-31",
                         lifestage.overlap = "always", verbose = FALSE)
```

## ✨ New lifestages {auto-animate="true"}

🛑 Lifestage `immigrant` no longer exists 🛑

## ✨ New lifestages {auto-animate="true"}

::: incremental
-   `founder_male`: sexually active adult male present in the crater at the start of the study (i.e. not cub, subadult, or natal).

-   `foreigner_X`: individual whose birth clan is unknown (X) or was born outside of the main clans (rim clans e.g. U, C) and that has made X observed selections. The number of previous selection events that occurred before the first selection is unknown, therefore X is a minimum number of previous selections.
:::

## ✨ New lifestages {auto-animate="true"}

::: incremental
-   `sexually_active`: Individuals that are sexually active (= "philopatric" + "disperser" + "selector_X" + 'foreigner_X' + "founder_male").

-   `selector`: Individuals born in the main clans that are sexually active (= "philopatric" + "disperser" + "selector_X").

-   `foreigner`: Individuals born outside the main clans that are sexually active (= 'foreigner_X').

-   `native`: Individuals that have not left their birth clan (= "cub" + "subadult" + "natal" + "philopatric").
:::

## ✨ New lifestages {auto-animate="true"}

::: callout-caution
### WARNING

`selector` meta-lifestage now excludes foreigners. Use `sexually_active` if you want *all* individuals that are sexually active.
:::

## ⚠️ Change default behaviour {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1-2"
# OLD BEHAVIOUR: Return all individuals ever BORN in Airstrip
create_id_starting.table.historic(clan = "A")
```

## ⚠️ Change default behaviour {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1-2"
# NEW BEHAVIOUR: Return all individuals ever PRESENT in Airstrip
create_id_starting.table(clan = "A")
```

## ⚠️ Change default behaviour {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1-3"
# NEW BEHAVIOUR: Return all individuals ever BORN in Airstrip
# using the clan.birth argument
create_id_starting.table(clan.birth = "A")
```

## The 'clan.birth' argument {auto-animate="true"}

Allows us to query current lifestage/clan *and* birth clan

```{r}
#| echo: true
#| code-line-numbers: "1-3|4|5-6"
# Individuals that were in Airstrip in 2010
# but that were not born there
create_id_starting.table(lifestage = "!native",
                         clan = "A",
                         from = "2010-01-01",
                         to = "2010-12-31")
```

## The 'clan.birth' argument {auto-animate="true"}

Allows us to query current lifestage/clan *and* birth clan

```{r}
#| echo: true
#| code-line-numbers: "1-2|7"
# Individuals that were in Airstrip in 2010
# but were born in Forest
create_id_starting.table(lifestage = "!native",
                         clan = "A",
                         from = "2010-01-01",
                         to = "2010-12-31",
                         clan.birth = "F")
```

## The 'clan.birth' argument {auto-animate="true"}

::: callout-caution
### WARNING

'clan.birth' does not consider from/to/at. For this we need to use lifestage = "cub".
:::

```{r}
#| echo: true
# Individuals born in Airstrip in 2010
create_id_starting.table(lifestage = "cub",
                         lifestage.overlap = "start",
                         clan = "A",
                         from = "2010-01-01",
                         to = "2010-12-31")
```

# 🌧️ Weather data ☀️

## Extract weather data {auto-animate="true"}

```{r}
#| echo: true
create_weather_starting.table(variable = c("temp", "rain"),
                              location = "acacia")
```

## Extract weather data {auto-animate="true"}

```{r}
#| echo: true
#| code-fold: true
raw_data <- create_weather_starting.table(variable = c("temp", "rain"),
                                          location = "acacia")

plot_data <- raw_data %>% 
  group_by(date = lubridate::as_date(date_time)) %>% 
  summarise(across(.cols = c(air_temp, precip), .fns = ~{mean(., na.rm = TRUE)}, .names = "{.col}_mean"),
            across(.cols = c(air_temp, precip), .fns = ~{min(., na.rm = TRUE)}, .names = "{.col}_min"),
            across(.cols = c(air_temp, precip), .fns = ~{max(., na.rm = TRUE)}, .names = "{.col}_max"))

ggplot(data = plot_data) +
  geom_ribbon(aes(x = date, ymin = air_temp_min, ymax = air_temp_max),
              fill = "grey70", alpha = 0.75) +
  geom_line(aes(x = date, y = air_temp_mean)) +
  labs(title = "Acacia air temperature",
       subtitle = "(Apr - Jul 2022)",
       y = "Mean air temperature (°C)") +
  scale_x_date(date_labels = "%B-%Y") +
  theme_classic() +
  theme(plot.title = element_text(face = "bold"),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major = element_line(colour = "grey10", size = 0.1))
```

## Extract weather data {auto-animate="true"}

```{r}
#| echo: true
fetch_weather_temp.mean(from = "2022-05-01", to = "2022-07-01",
                        location = "acacia")
```

::: fragment
```{r}
#| echo: true
fetch_weather_rain.max(from = "2022-05-01", to = "2022-07-01",
                       location = "acacia")
```
:::

## Extract weather data {auto-animate="true"}

::: callout-caution
### WARNING

Currently, only data from Acacia and Ngoitokitok are extractable with functions.
:::

::: callout-caution
### WARNING

On Windows, will need to use `download_package_csv(download.method = 'curl')` to download weather data.
:::

# [HOMEWORK:]{style="color:#ff2c2d;"} Combine what we've learnt!

## [TASK 1:]{style="color:#ff2c2d;"}

### Use `create_id_starting.table()` to find all individuals born in main clans in the years 1997 - 2021 (where we have full year observation).

<br>

::: fragment
### How many individuals have been born in total during this period?
:::

## [TASK 2:]{style="color:#ff2c2d;"}

### Extract individual birth clan, birth date, sex, and lifespan.

<br>

::: fragment
### How many individuals have missing data?
:::

<br>

::: fragment
### How would you interpet NAs in each column?
:::

## [TASK 3:]{style="color:#ff2c2d;"}

### Extract the *year* of birth for each individual.

<br>

::: fragment
### In which year were the most cubs born?
:::

<br>

::: fragment
### Is the most productive year the same for all clans?
:::

<br>

::: fragment
### BONUS: Use filter to return a data frame with only the best year(s) for each clan?
:::

## [TASK 4:]{style="color:#ff2c2d;"}

### Extract the *month* of birth for each individual.

<br>

::: fragment
### What was the most productive month in the crater from 1997-2021?
:::

<br>

::: fragment
### How many cubs were produced in that month?
:::
