---
title: "Data science with `{hyenaR}`:<br>LESSON 14"
format:
  revealjs:
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
library(stringr) ## Working with text
library(waldo) ## To compare objects
library(skimr) ## Inspect data
```

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

# GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}`

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### List of all adults of main clans with their highest lifetime rank.

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### [List of all adults of main clans]{style="color:#ff2c2d;"} with their highest lifetime rank.

## GOAL 1: 🧑‍🏫 Practical application of `{hyenaR}` {auto-animate="true"}

<br><br>

### [List of all adults of main clans]{style="color:#ff2c2d;"} [with their highest lifetime rank.]{style="color:#1b91ff;"}

## [List of all adults of main clans]{style="color:#ff2c2d;"} {auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "1-3|5-6|7-10|11-13"
## Work within the observation period...
start_date = find_pop_date.observation.first()
end_date = find_pop_date.observation.last()
create_id_starting.table(
  #Only known sex
  sex = c("female", "male"),
  #Individuals started and stopped being adults
  #in the study period (i.e. uncensored)
  lifestage = "adult",
  lifestage.overlap = "within",
  #Only use individuals born in main clans
  #Males that disperse from outside could have adult period unobserved
  clan.birth = find_clan_name.all(main.clans = TRUE),
  from = start_date,
  to = end_date
)
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "9-11"
create_id_starting.table(
  sex = c("female", "male"),
  lifestage = "adult",
  lifestage.overlap = "within",
  clan.birth = find_clan_name.all(main.clans = TRUE),
  from = start_date,
  to = end_date
) %>% 
  # Find the period during which ranks need to be calculated
  mutate(start_adult = fetch_id_date.at.age(ID, age = 2),
         end_adult = fetch_id_date.death(ID))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "1|8-9"
(expanded_df <- create_id_starting.table(
  sex = c("female", "male"),
  lifestage = "adult", lifestage.overlap = "within",
  clan.birth = find_clan_name.all(main.clans = TRUE),
  from = start_date, to = end_date) %>% 
  mutate(start_adult = fetch_id_date.at.age(ID, age = 2),
         end_adult = fetch_id_date.death(ID)) %>% 
  #Expand data to one row per 6 month period of an adults life
  reshape_row_date.seq(ID, from = start_adult, to = end_adult, by = "6 month"))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "2|3-4|5-7|8|9-10"
mainclans <- find_clan_name.all()
system.time({expanded_df_rank <- expanded_df %>% 
    # Extract current clan and per 6 months
    mutate(current_clan = fetch_id_clan.current(ID, at = date)) %>% 
    # Remove individuals that were outside the clan at any point
    group_by(ID) %>% 
    filter(all(current_clan %in% mainclans)) %>% 
    ungroup() %>% 
    #Extract ranks of remaining individuals
    mutate(rank = fetch_id_rank.std(ID, at = date))})
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

<br><br>

```{r}
#| echo: true
expanded_df_rank
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

```{r}
#| echo: true
#| fig-width: 14
#| fig-height: 7
#| code-fold: true
plot_data <- expanded_df_rank %>% 
  mutate(sex = fetch_id_sex(ID),
         philo = case_when(fetch_id_is.native(ID, at = date) ~ "philopatric",
                           TRUE ~ "disperser")) %>% 
  group_by(ID, philo) %>% 
  mutate(years_since_start = (date - min(date))/365,
         relative_rank_change = rank - min(rank))

ggplot(data = plot_data) +
  geom_line(aes(x = date, y = rank,
                group = ID, colour = ID)) +
  scale_x_date(breaks = seq(as.Date("1995-01-01"), as.Date("2020-01-01"), "5 year"),
               date_labels = "%Y",
               limits = c(as.Date("1995-01-01"), as.Date("2021-01-01"))) +
  labs(x = "Date", y = "Standardised rank") +
  facet_wrap(facets = ~sex + philo) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 10, colour = "black"))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

```{r}
#| echo: true
#| fig-width: 14
#| fig-height: 7
#| code-fold: true
plot_data2 <- plot_data %>% 
  filter(current_clan == "A")

ggplot(data = plot_data2) +
  geom_line(aes(x = date, y = rank,
                group = ID, colour = ID)) +
  scale_x_date(breaks = seq(as.Date("1995-01-01"), as.Date("2020-01-01"), "5 year"),
               date_labels = "%Y",
               limits = c(as.Date("1995-01-01"), as.Date("2021-01-01"))) +
  facet_wrap(facets = ~sex + philo) +
  labs(x = "Date", y = "Standardised rank",
       title = "Airstrip") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 10, colour = "black"),
        plot.title = element_text(size = 20, colour = "black", face = "bold"))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

```{r}
#| echo: true
#| fig-width: 14
#| fig-height: 7
#| code-fold: true
ggplot(data = plot_data2) +
  geom_line(aes(x = years_since_start, y = rank,
                group = ID, colour = ID)) +
  labs(y = "Standardised rank", x = "Years in lifestage",
       title = "Airstrip") +
  facet_wrap(facets = ~sex + philo) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 10, colour = "black"),
        plot.title = element_text(size = 20, colour = "black", face = "bold"))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

```{r}
#| echo: true
#| fig-width: 14
#| fig-height: 7
#| code-fold: true
ggplot(data = plot_data2) +
  geom_line(aes(x = years_since_start, y = rank,
                group = ID), colour = "grey60",
            alpha = 0.5) +
  geom_smooth(aes(x = years_since_start, y = rank),
              colour = "black", se = FALSE) +
  labs(y = "Standardised rank", x = "Years in lifestage",
       title = "Airstrip") +
  facet_wrap(facets = ~sex + philo) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 10, colour = "black"),
        plot.title = element_text(size = 20, colour = "black", face = "bold"))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

```{r}
#| echo: true
#| fig-width: 14
#| fig-height: 7
#| code-fold: true
plot_data3 <- plot_data %>% 
  filter(ID == "A-019")

ggplot(data = plot_data3) +
  geom_line(aes(x = date, y = rank,
                group = ID), size = 1,
            colour = "black") +
  geom_point(aes(x = date, y = rank),
             size = 2, shape = 21,
            colour = "black", fill = "white", stroke = 1) +
  labs(y = "Standardised rank", x = "Years in lifestage",
       title = "A-019") +
  scale_y_continuous(limits = c(NA, 1),
                     breaks = seq(-0.25, 1, 0.25)) +
  scale_x_date(breaks = seq(as.Date("1995-01-01"), as.Date("2020-01-01"), "5 year"),
               date_labels = "%Y",
               limits = c(as.Date("1995-01-01"), as.Date("2021-01-01"))) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 10, colour = "black"),
        plot.title = element_text(size = 20, colour = "black", face = "bold"))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

<br><br>

```{r}
#| echo: true
#| code-line-numbers: "2-4"
expanded_df_rank %>% 
  #Determine the max rank per ID
  group_by(ID) %>% 
  summarise(max_rank = max(rank)) %>% 
  arrange(desc(max_rank))
```

## [...with their highest lifetime rank.]{style="color:#1b91ff;"} {auto-animate="true"}

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 10
#| code-fold: true
plot_data4 <- expanded_df_rank %>% 
  #Determine the max rank per ID
  group_by(ID) %>% 
  summarise(max_rank = max(rank)) %>% 
  arrange(desc(max_rank)) %>% 
  mutate(birthclan = fetch_id_clan.birth(ID),
         sex = fetch_id_sex(ID))

ggplot(data = plot_data4) +
  geom_boxplot(aes(x = birthclan, y = max_rank, fill = sex)) +
  labs(x = "Clan", y = "Maximum rank") +
  theme_classic() +
  theme(axis.title = element_text(size = 15),
        axis.text = element_text(size = 10, colour = "black"),
        plot.title = element_text(size = 20, colour = "black", face = "bold"))
```
