---
title: "Data science with `{hyenaR}`:<br>LESSON 13"
format:
  revealjs:
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
#| code-line-numbers: "8"
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
library(stringr) ## Working with text
library(waldo) ## To compare objects
library(skimr) ## Inspect data
```

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

# GOAL 1: 📝[HOMEWORK:]{style="color:#ff2c2d;"} Lesson 10.

## [TASK 1:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Create a data frame of all females.

```{r}
#| echo: true
create_id_starting.table(sex = "female")
```

## [TASK 1:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Use `group_by()`/`summarise()` to show how many individuals are left or right censored.

```{r}
#| echo: true
#| code-line-numbers: "2-4|5-7"
create_id_starting.table(sex = "female") %>% 
  # Find which are left and right censored
  mutate(left_censor = fetch_id_is.censored.left(ID),
         right_censor = fetch_id_is.censored.right(ID)) %>% 
  #Group by and count
  group_by(left_censor, right_censor) %>% 
  count()
```

## [TASK 1:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Filter only un-censored individuals.

```{r}
#| echo: true
#| code-line-numbers: "2-3"
create_id_starting.table(sex = "female") %>% 
  # Filter uncensored individuals
  filter(!fetch_id_is.censored(ID))
```

## [TASK 2:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Extract lifetime reproductive success of females (consider only offspring that reach atleast 18 months old).

```{r}
#| echo: true
#| code-line-numbers: "3-4"
create_id_starting.table(sex = "female") %>% 
  filter(!fetch_id_is.censored(ID)) %>% 
  # Use age.mature and unit to find individuals
  mutate(LRS = fetch_id_number.offspring(ID, age.mature = 18, unit = "month"))
```

## [TASK 2:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Extract standardised rank within females at the date of adulthood (i.e. when they were 2 years old).

```{r}
#| echo: true
#| code-line-numbers: "4-7"
create_id_starting.table(sex = "female") %>% 
  filter(!fetch_id_is.censored(ID)) %>% 
  mutate(LRS = fetch_id_number.offspring(ID, age.mature = 18, unit = "month"),
         #Extract standardised rank within sexes
         #(i.e. just comparing females to other females)
         rank = fetch_id_rank.sex.std(ID,
                                      at = fetch_id_date.at.age(ID, age = 2)))
```

## [TASK 2:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Inspect the data. Are there any individuals you might exclude?

```{r}
#| echo: true
#| code-line-numbers: "6-8"
output <- create_id_starting.table(sex = "female") %>% 
  filter(!fetch_id_is.censored(ID)) %>% 
  mutate(LRS = fetch_id_number.offspring(ID, age.mature = 18, unit = "month"),
         rank = fetch_id_rank.sex.std(ID, at = fetch_id_date.at.age(ID, age = 2)))

#Check for NAs
any(is.na(output$LRS))
any(is.na(output$rank))
```

## [TASK 2:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Use `{skimr}` package to inspect data!

**NOTE** We need to look in RStudio because it doesn't work well on the slides!

```{r}
#| echo: true
#| eval: false
skim(output)
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### Create a new column with rank category (top rank \>0 or low ranking \<=0). [HINT:]{style="color:#ff2c2d;"} You might need to use `ifelse`.

```{r}
#| echo: true
#| code-line-numbers: "3-4"
output %>% 
  filter(!is.na(rank)) %>% 
  #Create high and low category
  mutate(rank_cat = ifelse(rank > 0, "high", "low"))
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### What is the mean and SE of lifetime reproductive success in each group? [HINT:]{style="color:#ff2c2d;"} Remember SE = $\frac{sd}{\sqrt N}$

```{r}
#| echo: true
#| code-line-numbers: "4-7"
output %>% 
  filter(!is.na(rank)) %>% 
  mutate(rank_cat = ifelse(rank > 0, "high", "low")) %>% 
  #Group-by summarise
  group_by(rank_cat) %>% 
  summarise(mean = mean(LRS),
            se = sd(LRS)/sqrt(n()))
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### [BONUS:]{style="color:#ff2c2d;"} Look at the help for the `case_when()` function. Can you use this to make 3 groups instead of 2?

```{r}
#| echo: true
#| code-line-numbers: "3-7|8-11"
output %>% 
  filter(!is.na(rank)) %>% 
  #Create 3 categories!
  mutate(rank_cat = ifelse(rank > 0.33, "rankgrp1",
                           ifelse(rank <= 0.33 & rank > -0.33,
                                  "rankgrp2",
                                  "rankgrp3"))) %>% 
  #Group-by/summarise
  group_by(rank_cat) %>% 
  summarise(mean = mean(LRS),
            se = sd(LRS)/sqrt(n()))
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### [BONUS:]{style="color:#ff2c2d;"} Look at the help for the `case_when()` function. Can you use this to make 3 groups instead of 2?

```{r}
#| echo: true
#| code-line-numbers: "3-6|7-10"
output %>% 
  filter(!is.na(rank)) %>% 
  #Do the same thing with case_when()
  mutate(rank_cat = case_when(rank > 0.33 ~ "rankgrp1",
                              rank <= 0.33 & rank > -0.33 ~ "rankgrp2",
                              rank <= -0.33 ~ "rankgrp3")) %>% 
  #group-by/summarise
  group_by(rank_cat) %>% 
  summarise(mean = mean(LRS),
            se = sd(LRS)/sqrt(n()))
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### [BONUS:]{style="color:#ff2c2d;"} Look at the help for the `case_when()` function. Can you use this to make 3 groups instead of 2?

```{r}
#| echo: true
#| code-line-numbers: "5-6"
output %>% 
  filter(!is.na(rank)) %>% 
  mutate(rank_cat = case_when(rank > 0.33 ~ "rankgrp1",
                              rank <= 0.33 & rank > -0.33 ~ "rankgrp2",
                              #Everything else that doesn't match others
                              TRUE ~ "rankgrp3")) %>% 
  group_by(rank_cat) %>% 
  summarise(mean = mean(LRS),
            se = sd(LRS)/sqrt(n()))
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### [BONUS:]{style="color:#ff2c2d;"} Look at the help for the `case_when()` function. Can you use this to make 3 groups instead of 2?

```{r}
#| echo: true
#| code-line-numbers: "1|4-5|6-8"
# Be careful using the TRUE group. It will include *anything* that doesn't match other criteria
output %>% 
  mutate(birthclan = fetch_id_clan.birth(ID),
         clan_grp = case_when(birthclan == "X" ~ "Outside",
                              TRUE ~ "Main")) %>% 
  filter(clan_grp == "Main") %>% 
  group_by(birthclan) %>% 
  count()
```

## [TASK 3:]{style="color:#ff2c2d;"}{auto-animate="true"}

<br>

### [BONUS:]{style="color:#ff2c2d;"} Look at the help for the `case_when()` function. Can you use this to make 3 groups instead of 2?

```{r}
#| echo: true
#| code-line-numbers: "1|4-6"
# Be careful *without* TRUE. Everything else is NA!
output %>% 
  mutate(birthclan = fetch_id_clan.birth(ID),
         clan_grp = case_when(birthclan %in% find_clan_name.all(main.clans = TRUE) ~ "Main")) %>% 
  group_by(clan_grp) %>% 
  count()
```

## [BONUS:]{style="color:#ff2c2d;"} Plotting{auto-animate="true"}

```{r}
#| echo: true
#| code-fold: true
#| fig-width: 5
#| fig-height: 6
library(ggdist)
plot_data <- output %>% 
  filter(!is.na(rank)) %>% 
  mutate(rank_cat = case_when(rank > 0.33 ~ "rankgrp1",
                              rank <= 0.33 & rank > -0.33 ~ "rankgrp2",
                              #Everything else that doesn't match others
                              TRUE ~ "rankgrp3"))

plot_data_summary <- plot_data %>% 
  group_by(rank_cat) %>% 
  summarise(mean = mean(LRS),
            se = sd(LRS)/sqrt(n()))

ggplot(data = plot_data_summary) +
  geom_col(aes(x = rank_cat, y = mean)) +
  geom_errorbar(aes(x = rank_cat, ymin = mean - se, ymax = mean + se), width = 0.2) +
  scale_x_discrete(labels = c("High", "Medium", "Low"), name = "") +
  scale_y_continuous(name = "Lifetime reproductive success (mean +- SE)") +
  theme_classic()
```
