---
title: "Data science with `{hyenaR}`:<br>LESSON 16"
format:
  revealjs:
    theme: [moon, css/custom.scss]
highlight-style: a11y
self-contained: true
self-contained-math: true
---

# Prepare our workspace

------------------------------------------------------------------------

<br>

### STEP 1: Load required packages

```{r}
#| echo: true
#| message: false
#| code-line-numbers: "9"
library(hyenaR) ## For our hyena specific functions
library(dplyr) ## For most data wrangling
library(ggplot2) ## For plotting
library(lubridate) ## Working with dates
library(tidyr) ## Extra data wrangling functions
library(stringr) ## Working with text
library(waldo) ## To compare objects
library(skimr) ## Inspect data
library(purrr) ## For loops in the tidyverse
```

------------------------------------------------------------------------

<br>

### STEP 2: Load the database

```{r}
#| echo: true
load_package_database.full(
  
  # Location of our database file
  db.path = "example_git/source_data/Fisidata_2022_08_10.sqlite"
  
)
```

# Today's goals

# GOAL 1: 🧑‍🏫 Using loops in R

# Adult sex ratio per year

## Without a loop {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "1-4|5-6|8-9"
#1997
# Extract all adults in 1997
ad_1997 <- create_id_starting.table(lifestage = "adult",
                                    from = "1997-01-01", to = "1997-12-31") %>% 
  #Extract sex information
  mutate(sex = fetch_id_sex(ID))

#Determine sex ratio
sexratio_1997 <- sum(ad_1997$sex %in% "male")/nrow(ad_1997)

sexratio_1997
```

## Without a loop {auto-animate="true"}

```{r}
#| echo: true
#| code-line-numbers: "8-14|16-17"
#1997
ad_1997 <- create_id_starting.table(lifestage = "adult",
                                    from = "1997-01-01", to = "1997-12-31") %>% 
  mutate(sex = fetch_id_sex(ID))

sexratio_1997 <- sum(ad_1997$sex %in% "male")/nrow(ad_1997)

### COPY-PASTE-EDIT THE SAME CODE FOR EVERY YEAR!
#1998
ad_1998 <- create_id_starting.table(lifestage = "adult",
                                    from = "1998-01-01", to = "1998-12-31") %>% 
  mutate(sex = fetch_id_sex(ID))

sexratio_1998 <- sum(ad_1998$sex %in% "male")/nrow(ad_1998)

tibble(year = 1997:1998,
       sexratio = c(sexratio_1997, sexratio_1998))
```

## Without a loop {auto-animate="true"}

<br><br>

### [RULE OF THREE:]{style="color:#ff2c2d;"} If you need to copy-paste the same code 3 or more times you need to rework (e.g. looping!)

## The 'for' loop {auto-animate="true"}

<br>

```{r}
#| echo: true
#| code-line-numbers: "1-3|5-6"
## For every value between 1997 and 2000...
## Create a variable called 'year' that has this value...
for (year in 1997:2000) {
  
  ## Run this code.
  print(paste("It's year:", year))
  
}
```

## The 'for' loop {auto-animate="true"}

<br>

### Why doesn't this do anything?!?!?

```{r}
#| echo: true
#| code-line-numbers: "1-3|5-8|9-14"
## For every value between 1997 and 2000...
## Create a variable called 'year' that has this value...
for (year in 1997:2000) {
  
  ## Start and end date of our code is now determined by
  ## the value of the variable 'year'
  start_date <- paste(year, "-01-01")
  end_date   <- paste(year, "-12-31")
  ad         <- create_id_starting.table(lifestage = "adult",
                                         from = start_date,
                                         to = end_date) %>% 
    mutate(sex = fetch_id_sex(ID))
  
  sum(ad$sex %in% "male")/nrow(ad)
  
}
```

## The 'for' loop {auto-animate="true"}

### Anything inside 'for' is lost unless we save it!

```{r}
#| echo: true
#| code-line-numbers: "1-2|16-17"
#Create an empty vector...
sex_ratio <- c()

## For every value between 1997 and 2000...
## Create a variable called 'year' that has this value...
for (year in 1997:2000) {
  
  ## Run this code.
  start_date <- paste(year, "-01-01")
  end_date   <- paste(year, "-12-31")
  ad         <- create_id_starting.table(lifestage = "adult",
                                         from = start_date,
                                         to = end_date) %>% 
    mutate(sex = fetch_id_sex(ID))
  
  #Append the new value to our vector
  sex_ratio <- append(sex_ratio, sum(ad$sex %in% "male")/nrow(ad))
  
}
```

## The 'for' loop {auto-animate="true"}

### Anything inside 'for' is lost unless we save it!

```{r}
#| echo: true
#| code-fold: true
#Create an empty vector...
sex_ratio <- c()

## For every value between 1997 and 2000...
## Create a variable called 'year' that has this value...
for (year in 1997:2000) {
  
  ## Run this code.
  start_date <- paste(year, "-01-01")
  end_date   <- paste(year, "-12-31")
  ad         <- create_id_starting.table(lifestage = "adult",
                                         from = start_date,
                                         to = end_date) %>% 
    mutate(sex = fetch_id_sex(ID))
  
  #Append the new value to our vector
  sex_ratio <- append(sex_ratio, sum(ad$sex %in% "male")/nrow(ad))
  
}

tibble(year = 1997:2000,
       sex_ratio = sex_ratio)
```

## 'apply' functions {auto-animate="true"}

### `sapply` returns a vector by default

```{r}
#| echo: true
#| code-line-numbers: "1|2-3|4-7"
sapply(
  # Loop through these values...
  X = 1997:2000,
  # Run this function
  FUN = function(year){
    paste("It's year", year)
  })
```

## 'apply' functions {auto-animate="true"}

### `sapply` returns a vector by default

```{r}
#| echo: true
#| code-line-numbers: "1|2-3|4-12"
sex_ratio <- sapply(
  ## Use each value 1997:2000...
  X = 1997:2000,
  ## To run this function...
  FUN = function(year){
    start_date <- paste(year, "-01-01")
    end_date   <- paste(year, "-12-31")
    ad         <- create_id_starting.table(lifestage = "adult", from = start_date, to = end_date) %>% 
      mutate(sex = fetch_id_sex(ID))
    
    #The output will be added to a vector
    sum(ad$sex %in% "male")/nrow(ad)
  })
```

## 'apply' functions {auto-animate="true"}

### `sapply` returns a vector by default

```{r}
#| echo: true
#| code-fold: true
sex_ratio <- sapply(
  ## Use each value 1997:2000...
  X = 1997:2000,
  ## To run this function...
  FUN = function(year){
    start_date <- paste(year, "-01-01")
    end_date   <- paste(year, "-12-31")
    ad         <- create_id_starting.table(lifestage = "adult", from = start_date, to = end_date) %>% 
      mutate(sex = fetch_id_sex(ID))
    
    #The output will be added to a vector
    sum(ad$sex %in% "male")/nrow(ad)
  })

tibble(year = 1997:2000,
       sex_ratio = sex_ratio)
```

## 'purrr' package {auto-animate="true"}

### `map` functions from package purrr can also be used to loop.

```{r}
#| echo: true
#| code-line-numbers: "1-2|3-9|11-13"
# Create a loop that already outputs a data frame (i.e. _df)!!
purrr::map_df(
  # Loop through each value of .x and run function .f
  .x = 1997:2000,
  .f = function(year){
    start_date <- paste(year, "-01-01")
    end_date   <- paste(year, "-12-31")
    ad         <- create_id_starting.table(lifestage = "adult", from = start_date, to = end_date) %>% 
      mutate(sex = fetch_id_sex(ID))
    
    #Output a data frame
    tibble(year = year,
           sex_ratio = sum(ad$sex %in% "male")/nrow(ad))
  })
```

## Work across a vector {auto-animate="true"}

### Many functions in `hyenaR` accept a vector!

```{r}
#| echo: true
#| code-line-numbers: "1-4|5-9|10-11"
# Create columns for start and end date...
tibble(year = 1997:2000) %>% 
  mutate(start_date = paste0(year, "-01-01"),
         end_date = paste0(year, "-12-31")) %>% 
  # Use fetch function that can take multiple start/end dates...
  mutate(nr_male = fetch_pop_number.male.adult(from = start_date,
                                               to = end_date),
         nr_ad = fetch_pop_number.anysex.adult(from = start_date,
                                               to = end_date),
         ## Calculate sex ratio
         sex_ratio = nr_male/nr_ad)
```

## Which is better? {auto-animate="true"}

<br>

- `hyenaR` can be a bit slower (for now), but should be more user friendly.

```{r}
#| eval: true
#| echo: false
print("purrr")
system.time({purrr::map_df(
  # Loop through each value of .x and run function .f
  .x = 1997:2000,
  .f = function(year){
    start_date <- paste(year, "-01-01")
    end_date   <- paste(year, "-12-31")
    ad         <- create_id_starting.table(lifestage = "adult", from = start_date, to = end_date) %>% 
      mutate(sex = fetch_id_sex(ID))
    
    #Output a data frame row
    tibble(year = year,
           sex_ratio = sum(ad$sex %in% "male")/nrow(ad))
  })})

print("sapply")
system.time({sex_ratio <- sapply(
  ## Use each value 1997:2000...
  X = 1997:2000,
  ## To run this function...
  FUN = function(year){
    start_date <- paste(year, "-01-01")
    end_date   <- paste(year, "-12-31")
    ad         <- create_id_starting.table(lifestage = "adult", from = start_date, to = end_date) %>% 
      mutate(sex = fetch_id_sex(ID))
    
    #The output will be added to a vector
    sum(ad$sex %in% "male")/nrow(ad)
  })

tibble(year = 1997:2000,
       sex_ratio = sex_ratio)})

print("for")
system.time({sex_ratio <- c()
  
  ## For every value between 1997 and 2000...
  ## Create a variable called 'year' that has this value...
  for (year in 1997:2000) {
    
    ## Run this code.
    start_date <- paste(year, "-01-01")
    end_date   <- paste(year, "-12-31")
    ad         <- create_id_starting.table(lifestage = "adult",
                                           from = start_date,
                                           to = end_date) %>% 
      mutate(sex = fetch_id_sex(ID))
    
    #Append the new value to our vector
    sex_ratio <- append(sex_ratio, sum(ad$sex %in% "male")/nrow(ad))
    
  }
  
  tibble(year = 1997:2000,
         sex_ratio = sex_ratio)})

print("hyenaR function")
system.time({tibble(year = 1997:2000) %>% 
  mutate(start_date = paste(year, "-01-01"),
         end_date = paste(year, "-12-31")) %>% 
  mutate(nr_male = fetch_pop_number.male.adult(from = start_date,
                                               to = end_date),
         nr_ad = fetch_pop_number.anysex.adult(from = start_date,
                                               to = end_date),
         sex_ratio = nr_male/nr_ad)})
```

## Which is better? {auto-animate="true"}

<br>

- `for` and `sapply` are base functions that require no dependency.

- `purrr` syntax works well for parallel processing packages.

